---
title: "Lab 1"
author: "Garrett Pinkston"
format: html
editor: visual
---

## Link to Github

github: <https://github.com/garrettpinkston2015/Computational-Methods>

## Problem 1- Wine Data

a\) Import Data and Rename Columns

```{r}

wine <- read.table('/Users/garrettpinkston/Desktop/Michigan/STAT506/Data/wine.data',sep=",")


wineNames <- c("Class", "Alcohol","Malic acid","Ash","Alcalinity of ash", "Magnesium", "Total phenols", 
  "Flavanoids","Nonflavanoid phenols", "Proanthocyanins", "Color intensity", "Hue", 
  "OD280/OD315 of diluted wines", "Proline")

colnames(wine) <- wineNames


```

b\) Verify wine counts match across each class

```{r}
counts <- c(59,71,48)
bool <- TRUE
for (i in 1:3){
  if(nrow(wine[wine$Class==i,]) != counts[i]){
    bool <- FALSE
  }
}

bool

```

c1\) Correlation between alcohol color and intensity?

```{r}
alcColorCor <- cor(wine$Alcohol,wine$`Color intensity`)
alcColorCor

```

c2\) Class with the highest and lowest correlation?

```{r}
corrs = c()
for (i in 1:3){
  wineCopy <- wine[wine$Class == i,]
  corrs[i] <- cor(wineCopy$Alcohol, wineCopy$`Color intensity`)
}

paste("Class", which.min(corrs), "achieved the lowest correlation with a coefficient of", round(min(corrs),4))
paste("Class", which.max(corrs), "achieved the highest correlation with a coefficient of", round(max(corrs),4))


```

c3\) What is the alcohol content of the wine with the highest color intensity?

```{r}
paste("The alcohol content of the wine with the highest color intensity is", wine$Alcohol[which.max(wine$`Color intensity`)])

```

c4\) What percentage of wines had a higher content of proanthocyanins compare to ash?

```{r}
higherPro <- nrow(wine[wine$Proanthocyanins > wine$Ash,])
paste0("Approximately ", round((higherPro/nrow(wine)*100),2), "% of the wines contain higher Proanthocyanins than Ash")


```

d\)  Create a table identifying the average value of each variable, providing one row for the overall average, and one row per class with class averages. (This table does not need to be “fancy” but should clearly identify what each value represents.)

```{r}
overallMeans <- colMeans(wine[-1], na.rm = TRUE)

wineSplit <- split(wine[-1],wine$Class)
classMeans <- sapply(wineSplit, function(x) colMeans(x,na.rm = TRUE))
#https://stackoverflow.com/questions/18879582/in-r-how-to-subset-data-frame-in-sapply

allMeans <- rbind(Overall = overallMeans, t(classMeans))

as.table(t(t(allMeans)))
```

e\)  Carry out a series of t-tests to examine whether the level of phenols differs across the three classes. Present the R output and interpret the results. (You may use an existing R function to carry out the t-test, or for minor extra credit, manually write your own calculation of the t-test p-values.)

```{r}

#take phenols of each class
phenolsClass_1 <- wine$`Total phenols`[wine$Class == 1]
phenolsClass_2 <- wine$`Total phenols`[wine$Class == 2]
phenolsClass_3 <- wine$`Total phenols`[wine$Class == 3]

#perform pairwise t-tests (1v2, 1v3, 2v3)
ttest_1_2 <- t.test(phenolsClass_1, phenolsClass_2)
ttest_1_3 <- t.test(phenolsClass_1, phenolsClass_3)
ttest_2_3 <- t.test(phenolsClass_2, phenolsClass_3)

paste0("The t-test between classes 1 and 2 was ", ttest_1_2$p.value,4)
paste0("The t-test between classes 1 and 3 was ", ttest_1_3$p.value)
paste0("The t-test between classes 2 and 3 was ", ttest_2_3$p.value)


```

With this output of p-values, we can reject the null hypothesis across all cases. All results are statistically significant, meaning that Total_phenols is an important variable to distinguish between pairs of classes.

## Problem 2- Askamanager.org Data

a\) Import Data and Rename Columns

```{r}

wine <- read.table('/Users/garrettpinkston/Desktop/Michigan/STAT506/Data/wine.data',sep=",")


wineNames <- c("Class", "Alcohol","Malic acid","Ash","Alcalinity of ash", "Magnesium", "Total phenols", 
  "Flavanoids","Nonflavanoid phenols", "Proanthocyanins", "Color intensity", "Hue", 
  "OD280/OD315 of diluted wines", "Proline")

colnames(wine) <- wineNames


```

## Problem 1- Wine Data

a\) Import Data and Rename Columns

```{r}

wine <- read.table('/Users/garrettpinkston/Desktop/Michigan/STAT506/Data/wine.data',sep=",")


wineNames <- c("Class", "Alcohol","Malic acid","Ash","Alcalinity of ash", "Magnesium", "Total phenols", 
  "Flavanoids","Nonflavanoid phenols", "Proanthocyanins", "Color intensity", "Hue", 
  "OD280/OD315 of diluted wines", "Proline")

colnames(wine) <- wineNames


```


