---
title: "HW2"
author: "Garrett Pinkston"
format: html
editor: visual
---

## **Problem 1 - Dice Game**

a\) Build the following versions

-   Version 1: Implement this game using a loop.

```{r}

loopDiceGame <- function(nrolls){
  wallet <- 0
  for (i in 1:nrolls) {
    roll <- sample(1:6,1)
    if ((roll == 3) || (roll == 5)) {
    wallet <- wallet + (2 * roll)
    }
    wallet <- wallet - 2
  }
  return(wallet)
}

set.seed(123)
loopDiceGame(1200)
```

-   Version 2: Implement this game using built-in R vectorized functions.

```{r}

vectorizedDiceGame <- function(nrolls){
  wallet <- 0
  
  allRolls <- sample(1:6,nrolls,replace=TRUE)
  
  wallet <- (allRolls == 3 | allRolls == 5) * 2 *allRolls

  wallet <- sum(wallet) - (2*nrolls)
  
  return(wallet)
}

set.seed(123)
vectorizedDiceGame(1200)
```

-   Version 3: Implement this by rolling all the dice into one and collapsing the die rolls into a single `table()`. (Hint: Be careful indexing the table - what happens if you make a table of a single dice roll? You may need to look to other resources for how to solve this.)

-   Version 4: Implement this game by using one of the “`apply`” functions.

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
